Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2014-06-29 pituba/article/article_spe.tex
abstract2

As mentioned, the equation developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, phase or thermal equilibrium. An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}. The novelty on hydrodynamics was the introduction of a macroscopic equation of motion using Darcy's law.

@@

The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system. For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo gas component can be defined. Similarly, a pseudo oil component can also be defined by subtracting all the pseudo gas component from oil phase composition. After defining these two pseudo-components, the number of transport equations needed for hydrocarbon species is reduced from a usually large number to only two.

@@

A great number of applications of material balance equation have been reported in the literature, but few of them explore its connection with black-oil equations. \cite{ertekin2001basic} briefly addressed this subject arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \vec{g}$) were neglected and if PVT properties were averaged in reservoir volume. The intuitive notion is that material balance equation is obtained if fluid motion is neglected and black-oil equations are integrated in reservoir volume. This work develops this process examining underlying details.

@@

The same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}, is also applied here to material balance. This approach extends material balance equation to saturated reservoirs under pressure maintenance (e.g. fluid injection). The applicability, however, depends on how important non-equilibrium effects are to phase equilibrium.

@@

After presenting material balance equation, sufficient conditions are shown for which material balance equation has only one solution for pressure. These conditions require fluids to be compressible and that gas dissolution occurs with reduction of the overall system volume. A simple scheme for numerical solution is also proposed based on classical root-finding algorithms for non-linear equations.

@@

This section starts presenting black-oil equations and it proceeds averaging quantities in space and examining simplifying assumptions for which material balance equation is obtained.

@@

We assume that the reservoir is a connected volume where hydrocarbon saturation is nonzero. Reservoir fluids are composed of oil and gas pseudo-components and water component. There can be three fluid phases: oil, gas and (liquid) water. Oil phase may be composed of oil and gas components, gas phase is composed uniquely by gas component and water phase uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases, i.e., gas phase composition is always that of a dry gas.

@@

Mass balance equations for each component read:

@@

Initial conditions are of no fluid motion and boundary conditions are of no normal flow at all boundaries except for interfaces with wells or aquifers.

@@

Temperature is a constant prescribed value and thus no equation for energy balance is needed. It follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if component number is restricted to the two pseudo-components. As temperature is assumed constant, oil phase density is a function of pressure alone. The same is valid for mass fractions of gas and oil components in oil phase:

@@

Porosity ($\phi$) is assumed a function of pore pressure ($p_P$). Because three phases are present in general, such quantity is defined by some relation of type \citep{kim2011rigorous}:
\begin{equation}
p_P=p_P\left(p_o,p_w,p_g,S_o,S_w,S_g\right) \, .
\end{equation}

@@

All quantities are initially functions of space and time and they become functions of time only after being averaged in space:

@@

Omitting function arguments:
\begin{equation}
\bar{S}_\alpha\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}S_\alpha \phi dV \, ,
\end{equation}

@@

We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases. In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to the volume at surface condition, i.e.:

@@

In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between volumes of dissolved gas and oil both measured in surface condition:

@@

An equation for bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure $R_s(p=P_b)$:
\begin{equation}
\begin{split}
R_{s2}\left(p=P_b\right) = &\Big[N\left(m B_{1}/B_{g1} +R_{s1}\right)\\
&+G_i-G_p\Big]/\left(N-N_p\right) \, .
\end{split}
\end{equation}

@@

It is now possible to obtain the so-called generalized material balance equation substituting the equations for phase saturation, Eqs.~\eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5}, into Eq.~\eqref{eq: S15}:

@@

\subsubsection{Relationship between the material balance equation and reservoir pressure}

@@

To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume hydrostatic equilibrium: $\vec{u}_\alpha \equiv 0$. This implies that $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, as a consequence, pressure becomes a function of time and depth only given by
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha^{ref} (t)}^{p_\alpha (t,z)} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'}\right)} = g\left(z - z_{ref}\right) \, ,
\end{equation}
where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha^{ref} (t)$ is the pressure at some reference depth $z_{ref}$.

It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly. Furthermore, it can be shown that neglecting the variation of fluid properties with depth does not introduce any important error. As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. For instance, the average oil density can be approximated to the local density at the reference level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$; the same applies to the formation volume factor: $B_o \approx \rho_o (t, z_{ref})/ \rho_o^{ST}$.

@@

Similarly for the gas phase, using the state equation $pV = Zn R T$, we obtain its isothermal compressibility:
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial p_g}\right)_{T} = \frac{1}{p_g}-\frac{1}{Z}\frac{\partial Z}{\partial p_g} \, .
\end{equation}
Neglecting the variation of Z-factor and using hydrostatic equilibrium again, we to obtain the derivative of gas density with depth as:
\begin{equation}\label{eq: gradgdz}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
\end{equation}

For investigating the orders of magnitude of Eqs.~\eqref{eq: gradodz} and \eqref{eq: gradgdz}, we substitute in them the values of $\rho_o = 10^3\ kg/m^3$ and $c_o = 10^{-9}\ Pa^{-1}$ for the oil phase, and $\mathcal{M}=19\ kg/kmol$ and $T=400\ K$ for the gas phase: 
\begin{equation}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z}\right)_{T,Y_o} \approx 1\times10^{-5}\ m^{-1} \, ,
\end{equation}

@@

Oil and gas densities are thus approximately constant with depth.
The reference depth for computing phase density (and also the formation volume factor) can be any arbitrary depth inside phase volume. Under the approximation of a constant density, it is easily seen that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$ where $z^0_\alpha$ is the depth of phase gravity center and $\alpha=w, o, g$.

For the case of an oil reservoir with a gas cap, the reference depth may be defined, with advantage, at the gas-oil contact. The point here is that at phase contact both gas and oil densities can be computed using the same pressure, i.e., the corresponding local value.

If the reference depth for oil phase is otherwise defined below gas-oil contact, gas density must be computed at some different depth still inside gas phase, i.e., at or above gas-oil contact. This is to avoid the effect of oil hydrostatic gradient in the computation of gas density,
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g  \, ,
\end{equation}
what is $\rho_o / \rho_g$ times greater than the sole effect of gas hydrostatic pressure on its density.

@@

Eq.~\eqref{eq: PVTH1} also provides a condition to check whether PVT data obtained from laboratory or empirical correlations is physically correct. It can be alternatively formulated as shown bellow in Eq.~\eqref{eq: vaporization} and it simply indicates that any vaporization must be followed by an increase in oil phase density:
\begin{equation}\label{eq: vaporization}
\frac{1}{\rho_o}\frac{d\rho_o}{dY_o} \geq Y_o \left( \frac{\rho_o}{\rho_g} - 1 \right) \, .
\end{equation}

>>>>

The equation developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, phase or thermal equilibrium. An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}. The novelty on hydrodynamics was the introduction of a macroscopic equation of motion using Darcy's law.

@@

The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system. For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo gas component can thus be defined. Similarly, a pseudo oil component can also be defined by subtracting all the pseudo gas component from oil phase composition. After defining these two pseudo-components, the number of transport equations needed for hydrocarbon components is reduced from a usually large number to only two.

@@

A great number of applications of material balance equation have been reported in the literature, but few of them explore its connection with black-oil equations. \cite{ertekin2001basic} briefly addressed this subject arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \vec{g}$) were neglected and if PVT properties were averaged in reservoir volume. This work develops this process examining underlying details.

@@

 The same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}, is also applied here to material balance. This approach extends the material balance equation to saturated reservoirs under pressure maintenance (e.g. fluid injection). The applicability, however, depends on how important non-equilibrium effects are to phase equilibrium.

@@

After presenting material balance equation, sufficient conditions are shown for which material balance equation has only one solution for pressure. These conditions require fluids to be compressible and that gas vaporization occurs with reduction of oil phase volume. A simple scheme for numerical solution is also proposed based on classical root-finding algorithms for non-linear equations.

@@

This section presents black-oil equations, performs averaging of quantities in space and examines simplifying assumptions necessary to obtain the material balance equation.

@@

We assume that the reservoir is a connected volume where hydrocarbon saturation is nonzero. Reservoir fluids are composed of oil and gas pseudo-components and water component. There can be three fluid phases: oil, gas and (liquid) water. Oil phase is composed of oil and gas components, gas phase is composed uniquely by gas component and, water phase, uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases. Gas phase composition is always that of a dry gas.

@@

Mass balance equation for each component reads:

@@

Initial conditions are of no fluid motion. Boundary conditions are of no normal flow at all boundaries except for interfaces with wells or aquifers.

@@

Temperature is a constant prescribed value and thus no equation for energy balance is needed. It follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if the component number is restricted to the two pseudo-components. As temperature is assumed constant, oil phase density is a function of pressure alone. The same is valid for mass fractions of gas and oil components in oil phase:

@@

Porosity ($\phi$) is assumed a function of pore pressure ($p_P$). Because three phases are present in general, such quantity is defined by some relation of the type \citep{kim2011rigorous}:
\begin{equation}
p_P=p_P\left(p_o,p_w,p_g,S_o,S_w,S_g\right) \, .
\end{equation}

@@

All quantities are initially functions of space and time, but they become functions of time only after being averaged in space:

@@

We proceed omitting function arguments:
\begin{equation}
\bar{S}_\alpha\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}S_\alpha \phi dV \, ,
\end{equation}

@@

We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases. In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to its volume at surface condition, i.e.:

@@

In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between the volume of dissolved gas and the volume of oil both measured in surface condition:

@@

An equation for the bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure, or $R_s(p=P_b)$:
\begin{equation}
\begin{split}
R_{s2}\left(p=P_b\right) = &\Big[N\left(m B_{1}/B_{g1} +R_{s1}\right)\\
&+G_i-G_p\Big]/\left(N-N_p\right) \, .
\end{split}
\end{equation}

@@

It is now possible to obtain the so-called generalized material balance equation by substituting the equations for phase saturation, Eqs.~\eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5}, into Eq.~\eqref{eq: S15}:

@@

\subsubsection{Relationship between the material balance equation and the reservoir pressure}

@@

To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume hydrostatic equilibrium: $\vec{u}_\alpha \equiv 0$. This implies that $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, as a consequence, pressure becomes a function of time and depth only given by the following expression:
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha^{ref} (t)}^{p_\alpha (t,z)} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'}\right)} = g\left(z - z_{ref}\right) \, ,
\end{equation}
where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha^{ref} (t)$ is its pressure at some reference depth $z_{ref}$.
It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly.

Furthermore, it can be shown that neglecting the variation of fluid properties with depth does not introduce any important error. As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. For instance, the average oil density can be approximated to the local density at the reference level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$; the same applies to the formation volume factor: $B_o \approx \rho_o (t, z_{ref})/ \rho_o^{ST}$.

@@

Similarly for the gas phase, using the state equation $pV = Zn R T$, we obtain its isothermal compressibility:
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial p_g}\right)_{T} = \frac{1}{p_g}-\frac{1}{Z}\frac{\partial Z}{\partial p_g} \, .
\end{equation}
Neglecting the variation of the Z-factor and using hydrostatic equilibrium again, we obtain the derivative of gas density with depth as:
\begin{equation}\label{eq: gradgdz}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
\end{equation}

For investigating the order of magnitude of Eqs.~\eqref{eq: gradodz} and \eqref{eq: gradgdz}, we substitute in them the values of $\rho_o = 10^3\ kg/m^3$ and $c_o = 10^{-9}\ Pa^{-1}$ for the oil phase, and $\mathcal{M}=19\ kg/kmol$ and $T=400\ K$ for the gas phase: 
\begin{equation}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z}\right)_{T,Y_o} \approx 1\times10^{-5}\ m^{-1} \, ,
\end{equation}

@@

Oil and gas densities are thus approximately constant with depth.
The reference depth for computing phase density (and also the formation volume factor) can be any arbitrary depth inside phase volume. Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$ where $z^0_\alpha$ is the depth of phase gravity center and $\alpha=w, o, g$. The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$, without fluid motion, and only equals the average reservoir pressure $\bar{p}_\alpha$ if $z^{ref}=z^0$.

For the case of an oil reservoir with a gas cap, the reference depth can be defined, with advantage, at the gas-oil contact. The point here is that at phase contact both gas and oil densities are computed using the same pressure, i.e., the corresponding local value.

If the reference depth for oil phase is otherwise defined below gas-oil contact, gas density must be computed at some different depth still inside gas phase, i.e., at or above gas-oil contact. This is to avoid the effect of oil hydrostatic gradient in the computation of gas density,
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g  \, ,
\end{equation}
what is $\rho_o / \rho_g$ times greater than the sole effect of gas hydrostatic pressure that was computed in Eq.~\eqref{eq: gradgdz}.

@@

Eq.~\eqref{eq: PVTH1} also provides a condition to check whether PVT data obtained from laboratory or empirical correlations is physically correct. It can be alternatively formulated as shown bellow in Eq.~\eqref{eq: vaporization}. It simply indicates that any vaporization must be followed by an increase in oil phase density:
\begin{equation}\label{eq: vaporization}
\frac{1}{\rho_o}\frac{d\rho_o}{dY_o} \geq Y_o \left( \frac{\rho_o}{\rho_g} - 1 \right) \, .
\end{equation}
